\large \bf Predicting Query Execution Time

### 2015-10-25 draft.tex
Revision #1

	\begin{abstract}
	The ability to estimate the query execution time is central for a number of tasks in database system
	such as query scheduling, progress monitoring and costing during query optimization. Recent work 
	has explored the use of statistical techniques in place of the manually constructed cost models used 
	in query optimization. Such techniques, which require as training data along with the 
	actual execution time, promises superior accuracies for they being able to account the for factors 
	such as hardware characteristics and bias in cardinality estimates. However, such techniques fail 
	to generalize i.e., produce poor estimates for queries that are not seen during the training.

	In this work, we propose and evaluate predictive modeling techniques that learn query 
	execution behavior at a fine grained operator level. For each operator, we consider different sets 
	of features and build different models for them. Since there are only finitely many operators in 
	database, this approach is practical and will be able to estimate any query as its a composition of
	many operators. We evaluate our approaches using TPC-H workload on PostgreSQL.

	\end{abstract}		
	\hfill 
	\begin{multicols}{2}
	\section{INTRODUCTION}
	Database systems can greatly benefit from accurate execution time predictions including: 
	\begin{itemize}
%	\item Query Optimizer: With accurate estimates, optimizer can rely on this metric to select the best plan among the many available plans.
	\item Admission control: Resource managers can use this metric to perform workload allocations such that the specific QoS are met \cite{activeSLA}.
	\item Query Scheduling: Knowing the execution time is crucial in deadline and latency aware 			scheduling
	\item Progress monitoring: Knowing the execution time of an incoming query can help avoid \textit{rogue queries} that are submitted in error and take an unreasonably long time to execute \cite{progress}.
	\end{itemize}
	Currently, execution time estimation is based on manually constructed
	models, which are part of the query optimizer and typically use
	combinations of weighted estimates of the number of tuples flowing
	through operators, column widths, etc. Unfortunately, such
	models often fail to capture several factors that affect the actual
	resource consumption. For example, they may not include detailed
	modeling of all of the various improvements made to database query
	processing – such as nested loop optimizations \cite{ganapathi, adaptive} which localize
	references in the inner subtree, and introduce “partially blocking”
	batch sorts on the outer side, thereby increasing the memory
	requirements and CPU time and reducing I/O compared to the traditional
	iterator model. Similarly, they may not accurately reflect
	specific hardware characteristics of the current production system
	or the impact of cardinality estimation errors. Analytical cost models predominantly 
	used by the current generation of query optimizers cannot
	capture these interactions and complexity; in fact, they are not designed to do so. 
	While they do a good job of comparing the costs of alternative query plans,
	they are poor predictors of plan execution latency. 
	Recent work \cite{ganapathi} showed this result for TPC-DS \cite{TPCDS}, and 
	in this work we do the same for TPC-H \cite{TPCH} data and queries.

@@

	\section{Background : Model Based Prediction}	
	We use the term model to refer to any predictive function such as
	Linear Regression, Least Squares and Support Vector Machines.
	Training a model involves using historical data sets to determine the
	best model instance that explains the available data. For example,
	fitting a function to a time series may yield a specific polynomial
	instance that can be used to predict future values.
	Model training (or building) requires selecting (a) the feature at-
	tributes, a subset of all attributes in the data set, and (b) a predic
	tion model, e.g., Linear Regression and Support Vector Machines
	(SVMs), to be used for modeling. In general, we \textit{cannot} know
	which model type and feature set will produce the most accurate
	model for a given data set without building and testing multiple
	models. In some cases, a domain expert can manually specify the
	feature attributes. Alternatively, feature attributes can be learned
	automatically; however, given a set of n attributes, trying the power
	set is prohibitively expensive if n is not small or training is expensive 
	thereby requiring heuristic solutions.
	Most approaches rank the candidate attributes (often based on their
	correlation to the prediction attribute using metrics such as information 
	gain or correlation coefficients) and use this ranking to guide
	a heuristic search \cite{datamining} to identify most predictive attributes tested
	over a disjoint test data set. 
%	In this paper, we will use a similar For-
%	ward Feature Selection algorithm based on linear correlation coef-
%	ficients [4]. This algorithm essentially performs a best-first search
%	in the model space. It starts with building models using small num-
%	ber of features and iteratively builds more complex and accurate
%	models by using more features. The features are considered with
%	respect to their correlation with the target/prediction attribute. The
%	training data set may be sampled to speed up the process.

@@

	Hypothesis testing and confidence interval estimations are two common
techniques for determining predictive accuracy \cite{forecasting}. As mentioned, 
it is not possible to estimate a priori what model would be
most predictive for a given data set without training/testing it. In this work, we are using
disjoint training and testing datasets that are generated from qgen tool of TPC-H workload 		\cite{TPCH}. The mean relative error is used to estimate the accuracy of our prediction models.

@@

	In this section, we elaborate on the overview of the approach which consists of 2 phases. Like 
	most other machine learning approaches, the first phase is an off-line training phase and the 
	second is on-line testing phase.
	During the training phase, we construct number of different 
	models for each type of physical database operator(e.g.,  Sort-Merge Join, Nested Loop, Index Nested 		Loop,..). Each operator will be associated with so called 'Best-Fit' model which will be determined 
	based on the Operator model selection described in section 3.1. This model will be invoked for the
	purpose of estimating the execution time for a certain operator. Since there are only finite number of 	operators and only a couple of physical implementations for them, the space consumption is not 				excessive.

@@

TO-DO :  
2. Describe the concept of multiple models for a single operator.
3. How do we select the best one among them. (Mention that current experiments don't use this idea, only one a simple non-linear model is used for all models with 2 features)

Analytical approaches problems:

counting the numbers of pages read is not sufficient as the discrepancy between random and sequential I/O is tremendous.

@@

	In \cite{analytical}, author's have done an analytical cost modeling where in they compute CPU and IO time individually by obtaining the number of operations performed and number of pages(sequential \& random) accessed respectively. They have also proposed a calibration phase where the cost model can account for the hardware changes which is absent in default Postgres model.  In previous unpublished work by Phankuri, they have devised methods to compute random page cost which was left out in the work\cite{analytical}. While this kind of approach seems natural, there's inherent ambiguity in knowing the number of pages accessed.  Even after knowing all this, there's no certain way we can compute IO time as cost model can never know whether a page is sequentially accessed or randomly accessed which impacts final cost computed heavily as there is a magnitude of difference in access times of sequential and random pages.
	Finally, there has also been work on query progress indicators \cite{progress}. Query progress indicators provide estimations for the comple-
	tion degrees of running queries. Such studies assume that the work
	done by individual query operators are transparent, i.e., externally
	visible. While these studies are also related to query execution performance,
	they do not provide predictions for the execution time of
	queries.

@@

>>>>

	\begin{abstract}
	The ability to estimate the query execution time is central for a number of tasks in database system
	such as query scheduling, progress monitoring and costing during query optimization. Recent work 
	has explored the use of statistical techniques in place of the manually constructed cost models used 
	in query optimization. Such techniques, which require as training data along with the 
	actual execution time, promises superior accuracies for they being able to account for the factors 
	such as hardware characteristics and bias in cardinality estimates. However, such techniques fail 
	to generalize i.e., produce poor estimates for queries that are not seen during the training.

	In this work, we propose and evaluate predictive modeling techniques that learn query 
	execution behavior at a fine grained operator level. For each \textit{physical implementation} of an 			operator, we consider different sets of features and build multiple models for them. Out of all those 		models, only the model that fits best is used later for prediciton.Since there are only finitely many 		operators in database, this approach is practical and will be able to estimate any query as its a 			composition of many operators. We evaluate our approaches using TPC-H workload on PostgreSQL.

	\end{abstract}		
	\hfill 
	\begin{multicols}{2}
	\section{INTRODUCTION}
	Database systems can greatly benefit from accurate execution time predictions including: 
	\begin{itemize}
%	\item Query Optimizer: With accurate estimates, optimizer can rely on this metric to select the best plan among the many available plans.
	\item Admission control: Resource managers can use this metric to perform workload allocations such that the specific QoS are met \cite{activeSLA}.
	\item Query Scheduling: Knowing the execution time is crucial in deadline and latency aware 			scheduling
	\item Progress monitoring: Knowing the execution time of an incoming query can help avoid \textit{rogue queries} that are submitted in error and take an unreasonably long time to execute \cite{progress}.
	\end{itemize}
	Currently, execution time estimation is based on manually constructed
	models, which are part of the query optimizer and typically use
	combinations of weighted estimates of the number of tuples flowing
	through operators, column widths, etc. Unfortunately, such
	models often fail to capture several factors that affect the actual
	resource consumption. For example, they may not include detailed
	modeling of all of the various improvements made to database query
	processing – such as nested loop optimizations \cite{nestedloops}. 
	Similarly, they may not accurately reflect
	specific hardware characteristics of the current production system
	or the impact of cardinality estimation errors. Analytical cost models predominantly 
	used by the current generation of query optimizers cannot
	capture these interactions and complexity; in fact, they are not designed to do so. 
	While they do a good job of comparing the costs of alternative query plans,
	they are poor predictors of plan execution latency. 
	Recent work \cite{ganapathi} showed this result for TPC-DS \cite{TPCDS}, and 
	in this work we do the same for TPC-H \cite{TPCH} data and queries.

@@

	\section{Background : Model Based Prediction}	
	We use the term model to refer to any predictive function such as
	Linear Regression, Least Squares and Support Vector Machines.
	Training a model involves using historical data sets to determine the
	best model instance that explains the available data. For example,
	fitting a function to a dataset may yield a specific linear/polynomial
	instance that can be used to predict future values.
	Model training (or building) requires selecting (a) the feature at-
	tributes, a subset of all attributes in the data set, and (b) a predic
	tion model, e.g., Linear Regression and Support Vector Machines
	(SVMs), to be used for modeling. In general, we \textit{cannot} know
	which model type and feature set will produce the most accurate
	model for a given data set without building and testing multiple
	models. In some cases, a domain expert can manually specify the
	feature attributes. 
%	Alternatively, feature attributes can be learned
%	automatically; however, given a set of n attributes, trying the power
%	set is prohibitively expensive if n is not small or training is expensive 
%	thereby requiring heuristic solutions.
%	Most approaches rank the candidate attributes (often based on their
%	correlation to the prediction attribute using metrics such as information 
%	gain or correlation coefficients) and use this ranking to guide
%	a heuristic search \cite{datamining} to identify most predictive attributes tested
%	over a disjoint test data set. 
%	In this paper, we will use a similar For-
%	ward Feature Selection algorithm based on linear correlation coef-
%	ficients [4]. This algorithm essentially performs a best-first search
%	in the model space. It starts with building models using small num-
%	ber of features and iteratively builds more complex and accurate
%	models by using more features. The features are considered with
%	respect to their correlation with the target/prediction attribute. The
%	training data set may be sampled to speed up the process.

@@

	Hypothesis testing and confidence interval estimations are two common
techniques for determining predictive accuracy \cite{forecasting}. As mentioned, 
it is not possible to estimate a priori what model would be
most predictive for a given data set without training/testing it. In this work, we are using
disjoint training and testing datasets that are generated from qgen tool of TPC-H workload 		\cite{TPCH}. Mean relative error(MRE) is used to estimate the accuracy of our prediction models.

Throughout the discussion of this work, we discriminate operator by its implementation method i.e., For example, implementations of Aggregate operator Group Aggregate, Hash Aggregate and Unique are all treated as different operators.

@@

	In this section, we elaborate on the overview of the approach which consists of 2 phases. Like 
	most other machine learning approaches, the first phase is an off-line training phase and the 
	second is on-line testing phase.
	During the training phase, we construct number of different 
	models for each type of database operator. Each operator will be associated with so called 'Best-Fit' 		model which will be determined 
	based on the model selection described in next section. Later, while in testing, only this model will 		be invoked for the purpose of estimating the execution time of an operator. Since there are 			only finite number of operators and only a couple of physical implementations for them, the space 			consumption is not excessive.

@@

@@

	In \cite{analytical}, author's have done an analytical cost modeling where in they compute CPU and IO time individually by obtaining the number of operations performed and number of pages(sequential \& random) accessed respectively. They have also proposed a calibration phase where the cost model can account for the hardware changes which is absent in default Postgres model.  In previous unpublished work by Phankuri, methods are provided to compute cost for reading random which was ambiguous the work\cite{analytical}. While this kind of approach seems natural, counting the numbers of pages read is not sufficient as the discrepancy between random and sequential I/O is tremendous. Unfortunately optimizer can never know whether a page is sequentially accessed or randomly accessed.

	Finally, there has also been work on query progress indicators \cite{progress}. Query progress indicators provide estimations for the comple-
	tion degrees of running queries. Such studies assume that the work
	done by individual query operators are transparent, i.e., externally
	visible. While these studies are also related to query execution performance,
	they do not provide predictions for the execution time of
	queries.

@@

	\bibitem{nestedloops}
	DeWitt, David J., Jeffrey F. Naughton, and Joseph Burger. "Nested loops revisited." Parallel and Distributed Information Systems, 1993., Proceedings of the Second International Conference on. IEEE, 1993.
